---
title: "Mastering Product Analytics - Final Assignment"
author: "Bence L. Toth (166504)"
date: "June 3rd, 2019"
output: html_document
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  error = FALSE,
  echo = TRUE,
  message = FALSE,
  fig.align = 'center',
  fig.height = 3.75,
  fig.width = 6
)
```

This analysis is based on the concepts of the book [Lean Analytics: Use Data Analytics to Build a Better Startup Faster](leananalyticsbook.com) by Alistair Croll and Benjamin Yoskovitz. It aims to carry out a simple data exploration regarding the topics of Acquisition, Activity and Retention, three key lifecycle stages of a product, according to the book.

```{r libraries, results = 'hide'}
rm(list = ls())
options(digits = 3)
options(scipen = 999)
library(dplyr)
library(tidyr)
library(ggplot2)
library(fpp2)
library(data.table)
library(kableExtra)
library(ggthemes)
custom_theme <- theme_tufte() +
  theme(legend.position = c(0.9, 0.9),
        legend.title = element_text(face = "italic", size = 12),
        axis.title = element_text(face = "bold", size = 14))

```

## About the data

```{r import}
data <- fread('data/activity.csv')
```

The example is based on a real Software as a Service, subscription based product. There are two datasets available:

* *registrations.csv*: unique users with basic demographics

* *activity.csv*: which users have been active in which month

More information about the datasets can be accessed on the project's [GitHub repository](https://github.com/BenceToth/ceu_product-analytics).

## Acquisition

Acquisition is a stage where you should generate attention through a variety of means, both organic and inorganic.
Relevant metrics to assess acquisition include traffic, mentions, cost per click, search results, cost of acquisition, open rate etc.

As a first step in the analysis, let's see the number of registrations in each month:

```{r acquisition_b1}
data %>%
  group_by(registration_month) %>% 
  summarize(count = n_distinct(id)) %>% 
  ggplot(aes(x = registration_month, y = count, fill = ifelse(registration_month <= 12, 1, 2))) +
  geom_col(show.legend = FALSE) +
  ggtitle('Figure 1: number of registrations per month') +
  custom_theme
```


```{r acquisition_b2, echo = FALSE}
regs <- data %>% 
  mutate(year = ifelse(registration_month <= 12, 1, 2)) %>% 
  group_by(year) %>% 
  summarize(regs = n_distinct(id))

drop <- data %>% 
  filter(registration_month %in% c(17, 18)) %>% 
  group_by(registration_month) %>% 
  summarize(regs = n_distinct(id))

drop <- round(drop[[1, 2]]/drop[[2, 2]] - 1, 4) * 100
```

We can see that year 2 has an overall lower number of registrations (`r regs[[1, 2]]` vs. `r regs[[2, 2]]`).
In the first year, most registrations occurred during the autumn/early winter period (September - November). We have no information on when this product was released to the market, the lower periods in the first half of year 1 might resemble the early adaptors, and the spike in the second part could be a successfull marketing campaign. After some lows during the cristmas holidays, there is a second peak until May of year 2, then a sudden, `r drop`% drop in registrations. This might be a failed product update, decrease in market reach or other reasons.

To dig deeper into the data, we can see the year-over-year growth of registrations:

```{r acquisition_t1a_1}
# add year and month
data <- data %>% 
  mutate(year = ifelse(registration_month <= 12, 1, 2),
         reg_month = ifelse(registration_month <= 12, registration_month, registration_month - 12))

# summarize registration number by year and month
yoyg <- data %>%   
  select(id, year, reg_month) %>% 
  group_by(reg_month, year) %>% 
  summarize(n_distinct(id)) %>% 
  rename(regs = 'n_distinct(id)') %>% 
  # get year-over-year growth
  spread(year, regs) %>% 
  rename(year_1 = '1',
         year_2 = '2') %>% 
  mutate(yoyg = year_2 / year_1) %>% 
  select(reg_month, yoyg) %>% 
  filter(reg_month <= 10)

yoyg$reg_month <- as.factor(yoyg$reg_month)

kable(yoyg, caption = 'Table 1: Year on Year Growth of registrations') %>%
  kable_styling('striped', full_width = F, position = 'center') %>%
  row_spec(0, color = 'white', background = 'blue', align ='c')
```

The next plot shows the growth rates for each month, together with the Naive forecast of Month 22 (October of the second year):

```{r acquisition_t1a_2}
ts <- ts(data = yoyg$yoyg, frequency = 1, start = 1, end = 9)
fc <- naive(ts, h = 1)

plot(fc, main = 'Figure 2: Year-on-year growth & Forecast from Naive method')
```

The year-over-year data seems to be decreasing, but keeping higher number of registrations in year 2 up until September. The Naive forecast for October is `r fc$mean[1]`.

We can investigate if there are any differences in registration numbers based on geographic regions:

```{r acquisition_t1b}
regions <- data %>%   
  select(id, year, reg_month, region) %>% 
  group_by(region, reg_month, year) %>% 
  summarize(n_distinct(id)) %>% 
  rename(regs = 'n_distinct(id)') %>% 
  # get year-over-year growth
  spread(year, regs) %>% 
  rename(year_1 = '1',
         year_2 = '2') %>% 
  mutate(yoyg = year_2 / year_1) %>% 
  select(region, reg_month, yoyg) %>% 
  filter(reg_month < 10)

ggplot(data = regions, aes(x = reg_month, y = yoyg, col = region)) +
  geom_line(size = 1.5) +
  ggtitle('Figure 3: Region differences in year-on-year growth') +
  geom_smooth(method = 'lm', se = FALSE, linetype = 'dashed') +
  custom_theme
```

Comparing the regions, there seems to be a regional change. *America* underperforms both EMEA and the rest of the world, except in the beginning of the year. All show a decreasing trend in registration numbers, but the *ROW* region seems to produce the highest growths. It is likely to drive future growth in registrations, while EMEA, and especially America might have different interest in the product, or they were less affected by the marketing campaigns.

## Activity

Activation is turning the resulting drive-by visitors into users who are somehow enrolled. It can be done by tracking a specific activity milestone, such as enrollments, signups, completed onboarding process, used the service at least once, subscriptions etc.

The plot below shows the number of active users in each month:

```{r activity_b2}
data %>%
  group_by(activity_month) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(x = activity_month, y = count, fill = ifelse(activity_month <= 12, 1, 2))) +
  geom_col(show.legend = FALSE) +
  ggtitle('Figure 4: number of active users per month') +
  custom_theme
```

The number of users increased until the summer of the first year, then it dropped, and only returned to the May level in September. It peaked during November, decreased during christmas and winter. The second year had its highest number of active users between March and May, and again decreased for the summer, showing a seasonality trait.

As we seen, the region of *America* struggles with new registrations. Let's see the percentage of *America* among active users in each month:

```{r activity_t2a}
data %>%
  group_by(region, activity_month) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(x = activity_month, y = count, fill = ifelse(region == 'America', 1, 2))) +
  geom_bar(position = 'fill', stat = 'identity', show.legend = FALSE) +
  ggtitle('Figure 5: number of active users per month') +
  custom_theme 
```

We can see that the number of users from *America* is on average between 17-22%. The company has to think about investing in other ways to drive the *American* market, or focus on the rest of their clients which seem to have better registration and activity rates.

```{r activity_t2b}

```

\# <add summary of plot here> \#

## Retention

The main task during the Retention phase is to convince users to come back repeatedly, exhibiting sticky behavior. It can be tracked by metrics such as engagement, time since last visit, daily and monthly active use, churns etc.

In the case of our example, we can calculate the second month retention rate (registered users in Month 1 who have been active in Month 2).

```{r retention_b1}
registered_month_1 <- data %>%
  filter(registration_month == 1) %>% 
  summarize(count = n_distinct(id))

activity_month_2 <- data %>% 
  filter(registration_month == 1 & activity_month == 2) %>% 
  summarize(count = n())

retention_rate_1_2 <- paste0(round(activity_month_2[[1]] / registered_month_1[[1]], 4) * 100, '%')
```

The second month retention rate is `r retention_rate_1_2`. It can be considered pretty good, it usually ranges from 20% to 60%.

Let's see the the same metric for year two (between Month 13 and 14).

```{r retention_b2}
registered_month_13 <- data %>%
  filter(registration_month == 13) %>% 
  summarize(count = n_distinct(id))

activity_month_14 <- data %>% 
  filter(registration_month == 13 & activity_month == 14) %>% 
  summarize(count = n())

retention_rate_13_14 <- paste0(round(activity_month_14[[1]] / registered_month_13[[1]], 4) * 100, '%')
```

`r retention_rate_13_14` of users who registered in January, second year are active in February. It is still an acceptable value, but what could be behind the 10% drop since last year?
It can be because the majority of users who are interested in the product have already signed up, so the pool of potential interested users for new registrations became smaller. This way the users who joined in the second year were not that engaged in the product and decided to opt out of it in the upcoming month with a higher percentage.


```{r retention_t3a}

```


```{r retention_t3b}

```

